\LARGE Untangling the link between traits, size and growth rate in plants

### 2016-10-13 ms/ms.tex
Incorporate suggestions from R Duursma throughout article

% Enable cross referencing to external documents
\usepackage{xr}
\externaldocument{ms-supp}
\externaldocument{figure_conceptual}
% Needed to set equation counter, following inclusion of figure 1
\setcounter{equation}{7}

@@

Plant species differ in many functional traits, including aspects of tissue construction and energy allocation, that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth, extending a widely-used theoretical model that links growth rate in seedlings with a single leaf trait to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Using a model based on energy uptake and allocation, we explain why the influence of prominent traits on growth rate often varies with plant size and why the impact of size varies among traits. Considering a plant's growth rate in height, we find the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) is present across sizes but is strongest at intermediate sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Our theoretical model shows how these changes in trait-growth relationships with size arise via changes in biomass partitioning and reproductive allocation. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results. By disentangling the effects of plant size, light environment and traits on growth rates, our results provide a solid theoretical foundation for trait ecology and thus provide a platform for understanding growth across diverse species around the world.

@@

In this article we outline how the growth of individual plants is influenced by various functional traits, also the plant's size and light environment. Functional traits capture core differences in the strategies plants use to generate and invest surplus energy \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nitrogen and water), species differ considerably in the rates at which resources are acquired and invested into different tissues. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers have looked to traits as potential indicators of plant growth and demography \citep{Poorter-2008, Wright-2010, Adler-2014}.\footnote{Add references from applied fields, such as invasive species, conservation.}

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly related to relative growth rate in plant mass \citep{Lambers-1992, Wright-2000}. leaf mass per unit leaf area and its correlate leaf lifespan were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, leaf mass per unit leaf area was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. So traits do not correlate simply and consistently to growth rate.

It is now becoming increasingly clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the strength of the correlation between some traits (including leaf mass per unit leaf area and maximum height) and growth rate was modified by size, while for other traits (including as wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, the direction of the correlations and their shifts with size supported predictions generated via a framework describing how traits influence growth.

The challenge of interpreting diverse empirical results seeking to link traits to growth rate was all the harder until the paper by \citet{Gibert-2016}, because we lacked any clear theoretical expectations on why the effect of traits may be moderated by size. Generating appropriate expectations is one of the primary roles for theory \citep{Kokko-2007}. A widely-used model for seedlings suggests a seedling's relative growth rate in mass is linearly and negatively related to the trait leaf mass per unit leaf area \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of the model suggests a similar relationship should hold at larger sizes \citep{Enquist-2007}. But as described above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers expected traits to linearly map onto growth rates, i.e. there is a ``fast'' and ``slow'' growth end for each trait \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One possible reason that theoretical predictions have not been supported is that the effects of traits in existing theory are realised mainly via influences on primary productivity (photosynthesis - respiration) or mass production \citep{Enquist-2007}. By contrast, the physiology of traits such as leaf mass per unit leaf area and wood density suggests they have influence by altering the allocation of energy among different tissues, rather than changing mass production \citep{Falster-2011, Gibert-2016}. A second concern for theory focussed on mass production is that altered mass production is only possible outcome. Measuring mass production is really only practical for small plants that can be easily harvested. For larger plants, growth is more-often measured as increment in either diameter or height. For example, research centres and forestry agencies have been recording stem diameter growth in 1000's of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. Theory considering only primary production is therefore of limited potential use.

In this paper we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth (Table \ref{tab:phenomena}); and thereby offer fresh insight into the way traits influence plant demography throughout across the life-cycle. This model can be viewed as an integration of several related approaches \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011}. Our primary focus in this article is to explain the pattern that has been gradually emerging -- that the effect of traits on plant growth may be modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. As noted above, the paper by \citet{Gibert-2016} outlined a framework describing how traits influence growth. Here we extend the results of \citet{Gibert-2016} to provide a functioning model and explain, from the point of view of energy uptake and allocation, why it makes sense that the effect of some traits changes with size. We also show how our approach can account for other phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment. Our view is that trait-based approaches -- which focus on differences among species -- should be integrated within a general model of plant growth, also able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Herault-2011}. In contrast, basal-area growth continues to increase with size \citep{Sillett-2010, Stephenson-2014}. And all growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Lusk-2008,Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Lusk-2008}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

@@

The {\plant} model builds on the widespread approach used in many vegetation models of explicitly modelling the amounts in different tissues within a plant \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, DeKauwe-2014} (Fig. \ref{fig:conceptual}a). We consider the masses $M_i$, areas $A_i$, and diameters $D_i$ of tissues, where the subscripts indicates tissue type: l=leaf, b=bark and phloem, s=sapwood, h=heartwood, r=root, a=live (l+b+s+r), t=total (l+b+s+h+r), and st=stem total (s+b+h). The total mass of living tissue is then $M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$, and the standing mass of the plant is $M_{\rm t}=M_{\rm a}+M_{\rm h}$. A summary of all variables, units and definitions is given in Table \ref{tab:definitions}-\ref{tab:traits}, with further details on the parameter values applied given in Table \ref{tab:params}.

We assume growth is fundamentally driven by energy income and subsequent distribution throughout the plant. Applying a standard approach, the amount of biomass available for growth, $\frac{{\rm d}B}{{\rm d}t}$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000} (Fig. \ref{fig:conceptual}a, eqn \ref{eq:dbdt}). Photosynthesis is the product of the average photosynthetic rate per unit leaf area ($\bar{p}$) and total leaf area ($A_{\rm l}$). The constant $y$ in eqn \ref{eq:dbdt} converts mol CO$_2$ into units of biomass. We assume that $\bar{p}$ increases with light but with diminishing returns, and respiration and turnover rates of different tissues are constants that might differ with traits (see below). While many vegetation models use a detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature, such detail is not needed here, as even with detailed models, $\frac{{\rm d}B}{{\rm d}t}$ always comes down to simple subtraction of income and losses.

@@

To model growth in either height, leaf area, basal area, standing mass or stem diameter then requires that we account not just for net mass production, but also for the costs of building new tissues, allocation to reproduction, and architectural layout. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016}. Relevant equations are given in Fig. \ref{fig:conceptual}b, (eqns \ref{eq:dbdt}-\ref{eq:dDdt}). As is evident in eqn  Fig. \ref{fig:conceptual}b, the growth rate in plant weight (eqn \ref{eq:dMtdt}), leaf area (eqn \ref{eq:daldt}), height (eqn \ref{eq:dhdt}), stem basal area (eqn \ref{eq:dast}), and stem diameter (eqn \ref{eq:dDdt}) all share some terms in common. Many of the terms in Fig. \ref{fig:conceptual} also vary intrinsically with size. The insets in Fig. \ref{fig:conceptual}b show the size-related patterns in different variables for a typical plant, obtained from applying the specific allocation model in the next section.

The growth rate of all size metrics in eqns \ref{eq:dMtdt}-\ref{eq:dDdt} depend on the product of biomass production $\frac{{\rm d}B} {{\rm d}t}$ (from eqn \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining  $1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$ fraction of mass produced is allocated to reproduction. Presumably, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, close to 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants. Note also that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even if the net amount of live mass $M_{\rm a}$ is not increasing.

The growth rate in the total standing mass of the plant, $\frac{{\rm d}M_{{\rm t}}}{{\rm d}t}$ is then the sum of sapwood turnover (which is replaced by an equivalent amount) and increment in live mass (eqn \ref{eq:dMtdt}).

@@

Eqn \ref{eq:dbdt} can also be used to estimate a measure of shade tolerance, sometimes referred to as the ``whole-plant-light-compensation point'' ({\wplcp}): the light level at which a plant's photosynthetic gains just balance the costs of tissue turnover and respiration \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. In general, average assimilation rate per leaf area $\bar{p}$ increases with light level or canopy openness, $E$. The {\wplcp} can be estimated by solving for the the value of canopy openness $E=E^*$ giving $\frac{{\rm d}B} {{\rm d}t}(E)=0$ (Fig. \ref{fig:wplcp_idea}). From eqn \ref{eq:dbdt}, this occurs when
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum_{i={\rm l,b,s,r}} \, M_{\rm i} \left(\frac{k_{\rm i}}{y} + r_{\rm i}\right)}{A_{\rm l}},
\end{equation}
where $\bar{p}(E)$ is the rate of photosynthesis at canopy openness $E$. The {\wplcp} occurs at the points where the photosynthetic production (per unit leaf area) line intersects with the sum of maintenance and respiration costs (per unit leaf area) for each tissue (Fig. \ref{fig:wplcp_idea}). Traits will influence the {\wplcp} if they effect either income or costs. Also, since the amount of stem support naturally increases with height, the {\wplcp} also naturally increases with height \citep{Givnish-1988} (Fig. \ref{fig:wplcp_idea}).

@@

It worth noting that because eqns \ref{eq:dMtdt}-\ref{eq:dDdt} are derived using the chain rule of differential calculus, they are mathematically true and must therefore hold for any potential growth model. To make explicit predictions via this framework, then requires an explicit model of plant construction and function; i.e. we must put equations behind all there terms in Fig. \ref{fig:conceptual}b.

Here we adopt a model of plant construction and function that can be considered a first-order functional-balance model, similar to those implemented by \citet{Makela-1997} and \citet{Moorcroft-2001}. We could also call it ``isometric'', because the assumptions preserve the proportions among the relevant variables as the plant grows. Table \ref{tab:allometry} provides key equations of the model (see \citet{Falster-2016} for full derivation). In particular, we assume:
\begin{description}
\item[Assumption 1:] A log-log scaling relationship with slope of 0.5 between the leaf area of a plant and its height (eqn \ref{eq:ha}).
\item[Assumptions 2:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of sapwood in a plant (eqn \ref{eq:as}).
\item[Assumptions 3:] A log-log scaling relationship with slope of 1.0 between the leaf area and cross-sectional area of bark and phloem in a plant (eqn \ref{eq:ab}).
\item[Assumptions 4:] A log-log scaling relationship with slope of 1.0 between the leaf area and mass of fine roots in a plant (eqn \ref{eq:mr}).
\item[Assumption 5:] The vertical distribution of leaf within the canopy of an individual plant remains constant through ontogeny.
\end{description}
Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are realisations of the pipe model \citep{Shinozaki-1964}, whereby the cross-sectional area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant (assumption 5), we use the model of \citet{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1 (for details, see Supplementary Materials).

@@

An important prediction arising from the eqns in Table \ref{tab:allometry} is that the amount of live stem tissue supporting each unit of leaf area increases linearly with  height:
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H,
\end{equation}
where $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are all traits (Table \ref{tab:traits}), i.e. properties that are assumed approximately constant through ontogeny.

@@

For the most part, parameters used in the current analysis were sourced from \citet{Falster-2016} (see Table \ref{tab:params} for values). The only exceptions are i))parameters effecting the relationships outlined in Table \ref{tab:allometry}a, which were fit using data described below, and ii) parameters describing the function for reproductive allocation. By default, we set $r_{\rm r1}=0.8$ and $r_{\rm r2}=10$, implying a relatively rapid transition to reproduction at $H=H_{{\rm mat}}$ (see panel for $\frac{{\rm d}M_{{\rm a}}}{{\rm d}B}$ in Fig. \ref{fig:conceptual}).

@@

\subsection{Relationship to a classic model for seedling growth}

>>>>

% Enable cross referencing to external documents
\usepackage{xr}
\externaldocument{ms-supp}
\externaldocument{figure_conceptual}

@@

Plant species differ in many functional traits that drive differences in rates of photosynthesis, biomass allocation, and tissue turnover. Yet, it remains unclear how -- and even if -- such traits influence whole-plant growth, with the simple linear relationships predicted by existing theory often lacking empirical support. Here we present a new theoretical framework for understanding the effect of diverse functional traits on plant growth and shade-tolerance, extending a widely-used theoretical model that links growth rate in seedlings with a single leaf trait to explicitly include influences of size, light environment, and five other prominent traits: seed mass, height at maturation, leaf mass per unit leaf area, leaf nitrogen per unit leaf area, and wood density. Based on biomass production and allocation, this framework explains why the influence of prominent traits on growth rate and shade tolerance often varies with plant size and why the impact of size on growth varies among traits. Considering  growth rate in height, we find the influence of: i) leaf mass per unit leaf area is strong in small plants but weakens with size, ii) leaf nitrogen per unit leaf area does not change with size, iii) wood density is present across sizes but is strongest at intermediate sizes, iv) height at maturation strengthens with size, and v) seed mass decreases with size. Moreover, we show how traits moderate plant responses to light environment and also determine shade tolerance, supporting diverse empirical results. By disentangling the effects of plant size, light environment and traits on growth rates, our results provide a solid theoretical foundation for trait ecology and thus provide a platform for understanding growth across diverse species around the world.

@@

% XX - The main comment I have is whether the strength of the 'correlation' between traits and growth rates can be summarized in some metric, so that it can be directly shown. To me Fig 4 is not very clear on showing how size affects the trait-growth correlations. For example, for wood density, I see a relation for all size classes. For the largest trees, it may look like a weaker correlation but isn't that because growth rates are lower? The same with Narea, to me they all look the same (except again, larger trees grow slower but it’s hard to tell there for me whether the trait-growth correlation is any different)
% XX -  The shade tolerance (WPLCP) analysis seems off-topic (it also not mentioned in the abstract). Can this be integrated more closely with the main objective, i.e. correlation between traits and growth rate and its dependence on size?

In this article we outline how the growth of individual plants is influenced by various functional traits, also the plant's size and light environment. Functional traits capture core differences in the strategies plants use to generate and invest surplus energy \citep{Westoby-2002, Wright-2004, Chave-2009}. Although most woody plants have the same basic physiological function and key resource requirements (carbon, nitrogen and water), species differ considerably in the rates at which resources are acquired and invested into different tissues. During the last two decades, trade-offs related to some prominent traits have been quantified, with values for traits such as leaf mass per unit leaf area, wood density, and seed size now available for up to 10\% of the world's 250000 plant species \citep{Cornwell-2014}. As data has accumulated, researchers are increasingly looking to traits to predict patterns in plant growth, demography, life history and performance \citep{Poorter-2008, Wright-2010, VanKleunen-2010, Adler-2014}.

While the influence of traits on elements of plant physiological function has been increasingly quantified and understood, attempts at using traits to predict demographic rates have met with mixed success \citep{Poorter-2006, Poorter-2008,Wright-2010,Herault-2011,Paine-2015}. In seedlings, the trait leaf mass per unit leaf area -- the central element of the leaf economics spectrum \citep{Wright-2004} -- was found to be tightly related to relative growth rate in plant mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}, as predicted from a mathematical model of growth rate (see below). Leaf mass per unit leaf area and its correlate leaf lifespan were also linked to height growth rate for small seedlings and saplings \citep{Reich-1992, Poorter-2006}. These early successes prompted researchers to search for similar relationships in large plants. However, the results showed that in saplings and trees, leaf mass per unit leaf area was not correlated with growth rate \citep{Poorter-2008, Wright-2010, Herault-2011, Paine-2015}. Meanwhile, other traits such as wood density showed strong relationships to growth in large plants \citep{Wright-2010,Herault-2011}, but less so in small plants \citep{Castro-1998}. So traits do not correlate simply and consistently to growth rate.

Recently, it has become clear that the effect of traits on plant growth can be modified by plant size \citep{Falster-2011, Ruger-2012, Iida-2014, Visser-2016, Gibert-2016}. A recent meta-analysis of 103 studies reporting \textgreater 500 correlations by \citet{Gibert-2016} provides the most compelling evidence. \citet{Gibert-2016} showed that the strength of the correlation between some traits (including leaf mass per unit leaf area and maximum height) and growth rate was modified by size, while for other traits (including as wood density and assimilation rate per leaf area) the sign of the correlation remained the same, irrespective of size. Importantly, the direction of the correlations and their shifts with size supported predictions generated via a framework describing how traits influence growth.

The challenge of interpreting diverse empirical results seeking to link traits to growth rate was all the harder because, until the paper by \citet{Gibert-2016}, we lacked any clear theoretical expectations on why the effect of traits may be moderated by size. Generating appropriate expectations is one of the primary roles for theory \citep{Kokko-2007}. A widely-used equation for seedlings suggests that, all else being equal, a seedling's relative growth rate in mass is linearly and negatively related to the trait leaf mass per unit leaf area \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. An extension of the model suggests a similar relationship should hold at larger sizes \citep{Enquist-2007}. But as noted above, the prediction for large plants is not supported by empirical results. Meanwhile, theoretical predictions on how other traits should influence growth are largely absent. Without further guidance, many researchers expected traits to linearly map onto growth rates, i.e. there is a ``fast'' and ``slow'' growth end for each trait \citep[e.g.][]{Grime-1977, Poorter-2008, Chave-2009, Paine-2015}.

One possible reason that theoretical predictions have not been supported is that the effects of traits in existing theory are realised mainly via influences on primary productivity (photosynthesis - respiration) or mass production \citep{Wright-2000,Enquist-2007}. By contrast, the physiology of traits such as leaf mass per unit leaf area and wood density suggests they have influence by altering the allocation of energy among different tissues, rather than changing mass production \citep{Falster-2011, Duursma-2016, Gibert-2016}. A second concern for theory focussed on mass production is that measuring mass production is really only practical for small plants that can be easily harvested. For larger plants, growth is more-often measured as increment in either diameter or height. For example, research centres and forestry agencies have been recording stem diameter growth in 1000's of plots across the globe \citep{Purves-2008, Anderson-2015, Kunstler-2016}. Theory considering only primary production therefore offers limited insight.

In this paper we show how a new mechanistic growth model -- called {\plant} \citep{Falster-2016} -- can explain diverse empirical phenomena, including a size-dependent effect of traits on growth and shade tolerance (Table \ref{tab:phenomena}); and thereby offer fresh insight into the way traits influence plant demography across the life-cycle. Broadly, the {\plant} model builds off several related approaches, focussing on production and allocation of biomass \citep[e.g.][]{Givnish-1988, Yokozawa-1995,Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011}. Our primary focus in this article is to explain the pattern that has been gradually emerging -- that the effect of traits on plant growth may be modified by plant size \citep{Ruger-2012, Iida-2014, Gibert-2016}. As noted above, the paper by \citet{Gibert-2016} outlined a framework describing how traits influence growth. Here we extend the results of \citet{Gibert-2016} to provide a functioning model and explain, from the point of view of energy uptake and allocation, why it makes sense that the effect of some traits changes with size. We also show how our approach can account for other phenomena, including changes in growth and shade tolerance with traits, individual size, and light environment. Our view is that trait-based approaches -- which focus on differences among species -- should be integrated within a general model of plant growth, also able to capture patterns of growth through ontogeny. Growth rates tend to show hump-shaped relationship with size, when expressed as either height \citep{Sillett-2010, King-2011} or diameter growth \citep{Herault-2011} or biomass production \citep{Givnish-1988, Koch-2004}. In contrast, the growth rate of plant mass continues to increase with size \citep{Sillett-2010, Stephenson-2014}. And all growth measures decrease sharply with size when expressed as relative growth rates \citep{Rees-2010, Iida-2014}. Shade tolerance also varies among species, correlates with traits \citep{Lusk-2008,Poorter-2006}, and tends to decrease with increasing size \citep{Givnish-1988, Lusk-2008}. These diverse phenomena -- summarised in Table \ref{tab:phenomena} -- are deserving of a comprehensive and integrated explanation.

@@

The {\plant} model builds on the widespread approach used in many vegetation models of explicitly modelling the amounts of biomass in different tissues within a plant \citep[e.g.][]{Givnish-1988, Makela-1997, Moorcroft-2001, Sitch-2008, Falster-2011, King-2011, DeKauwe-2014} (Fig. \ref{fig:conceptual}a). We consider the masses $M_i$, areas $A_i$, and diameters $D_i$ of tissues, where the subscripts indicates tissue type: l=leaf, b=bark and phloem, s=sapwood, h=heartwood, r=root, a=live (l+b+s+r), t=total (l+b+s+h+r), and st=stem total (s+b+h). The total mass of living tissue is then $M_{\rm a}=M_{\rm l}+M_{\rm b}+M_{\rm s}+M_{\rm r}$, and the standing mass of the plant is $M_{\rm t}=M_{\rm a}+M_{\rm h}$. A summary of all variables, units and definitions is given in Table \ref{tab:definitions}-\ref{tab:traits}, with further details on the parameter values applied given in Tables \ref{tab:params_core}--\ref{tab:params_hyper}.

We assume growth is fundamentally driven by biomass production and its subsequent distribution throughout the plant. Applying a standard approach, the amount of biomass available for growth, $\frac{{\rm d}B}{{\rm d}t}$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000}:
\begin{equation}\label{eq:dbdt}
\underbrace{\strut \frac{{\rm d}B}{{\rm d}t}}_\textrm{net biomass production}
= \underbrace{\strut \alpha_{\rm bio} }_\textrm{mass per C} \,
\underbrace{\strut \alpha_{\rm y} }_\textrm{yield}
\big( \underbrace{\strut A_{\rm l} \, \bar{p}(E)}_\textrm{photosynthesis} -
\underbrace{\strut \, \sum_{i = {\rm l}, {\rm b}, {\rm s}, {\rm r}}{M_i \, r_i}}_\textrm{respiration}\big)
- \underbrace{\strut \sum_{i = {\rm l}, {\rm b}, {\rm s},  {\rm r}}{M_i \, k_i}}_\textrm{turnover}.
\end{equation}
Photosynthesis is the product of the average photosynthetic rate per unit leaf area ($\bar{p}(E)$) and total leaf area ($A_{\rm l}$). We assume that $\bar{p}$ increases with canopy openness $E$, as per a standard light-response curve (Fig. \ref{fig:wplcp_idea}; for details, see Supplementary Materials), and respiration and turnover rates of different tissues are constants that might differ with traits (see below). The constants $\alpha_{\rm y}$ and $\alpha_{\rm bio}$ account for growth respiration and the conversion of CO$_2$ into units of biomass, respectively. While the {\plant} model can easily accommodate competitive shading via influences on $E$, in this analysis we grow individual plants under a fixed light environment, so that we can better understand the intrinsic trait- and size-related effects. Also, many vegetation models use a more-detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature. Such detail is not needed here because it will not change general model behaviour, even it would alter  he absolute values of predicted growth rates. Even with detailed models, $\frac{{\rm d}B}{{\rm d}t}$ always comes down to simple subtraction of income and losses.

\subsection{Original model for mass-based relative growth rate}

Earlier studies focussing on seedlings used a mathematical model to relate the trait leaf mass per area ($\phi$) to relative growth rate in mass \citep{Lambers-1992, Cornelissen-1996, Wright-2000}. In fact, that model is a special case of the more-extended approach described in the following sections and can be derived from eqn \ref{eq:dbdt} as follows. For seedlings, which are young and mostly leaf, we first ignore all turnover terms as well as the respiration terms for non-leaf tissues in eqn \ref{eq:dbdt}. Net production (from eqn \ref{eq:dbdt}) then becomes a linear function of leaf area, making relative growth rate in mass a linear function of $\phi$:
\begin{equation}\label{eq:RGR}
\frac{{\rm d}B}{{\rm d}t}\frac{1}{M_{\rm a}}  \approx P_{\rm net} \times \phi^{-1} \times \frac{M_{\rm l}}{M_{\rm a}}, \end{equation}
where $P_{\rm net} = \alpha_{\rm bio} \, \alpha_{\rm y}\,(\bar{p}(E) - r_{\rm l})$. Although eqn \ref{eq:RGR} captures patterns of growth in seedlings in relation to $\phi$ \citep{Wright-2000}, this approximation does not easily extend to the variables that are routinely collected for large trees: namely plant height ($H$), stem-basal area ($A_{\rm st}$) or stem diameter $D$. The derivations below makes these links clear.

@@

% Update equation counter, following inclusion of figure 1
\setcounter{equation}{9}

To model growth in either plant height ($H$), leaf area ($A_{\rm l}$), basal stem area ($ A_{\rm st}$), standing mass ($M_{\rm t}$) or stem diameter ($D$) then requires that we account not just for net mass production, but also for the costs of building new tissues, allocation to reproduction, and architectural layout. Mathematically, these growth rates can be decomposed into a product of physiologically relevant terms \citep{Falster-2011, Gibert-2016}. Relevant equations are given in Fig. \ref{fig:conceptual}b, (eqns \ref{eq:dMtdt}-\ref{eq:dDdt}). As is evident in eqn  Fig. \ref{fig:conceptual}b, the growth rate in plant weight ($\frac{\ud M_{\rm t} }{\ud t}$; eqn \ref{eq:dMtdt}), leaf area ($\frac{\ud A_{\rm l}}{\ud t}$; eqn \ref{eq:daldt}), height ($\frac{\ud H}{\ud t}$; eqn \ref{eq:dhdt}), stem basal area ($\frac{\ud  A_{\rm st}}{\ud t}$; eqn \ref{eq:dast}), and stem diameter ($\frac{\ud D}{\ud t}$; eqn \ref{eq:dDdt}) all share some terms in common. Many of the terms in Fig. \ref{fig:conceptual} also vary intrinsically with size. The insets in Fig. \ref{fig:conceptual}b show the size-related patterns in different variables for a typical plant, obtained from applying the specific allocation model in the next section.

The growth rate of all size metrics in eqns \ref{eq:dMtdt}-\ref{eq:dDdt} depend on the product of biomass production $\frac{{\rm d}B} {{\rm d}t}$ (from eqn \ref{eq:dbdt}) and the fraction of biomass allocated to growth of the plant, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$, which varies from 0-1. The remaining  $1- \frac{{\rm d}M_{\rm a}}{{\rm d}B}$ fraction of mass produced is allocated to reproduction. In plants, $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ starts high, 1.0 for seedlings, and then decreases through ontogeny, potentially to zero in fully mature plants \citep{Wenk-2015}. Note also that $\frac{{\rm d}M_{\rm a}}{{\rm d}B}$ is the allocation of biomass \emph{after} replacing parts lost due to turnover. So a plant with $\frac{{\rm d}M_{\rm a}}{{\rm d}B}=0$ will continue to produce some new leaves and increase in stem diameter, even if the net amount of live mass $M_{\rm a}$ is not increasing.

The growth rate in the total standing mass of the plant (eqn \ref{eq:dMtdt}) is then the sum of heartwood formation (=sapwood turnover) and any increment in live mass.

@@

Eqn \ref{eq:dbdt} can also be used to estimate a measure of shade tolerance, sometimes referred to as the ``whole-plant-light-compensation point'' ({\wplcp}): the light level at which a plant's photosynthetic gains just balance the costs of tissue turnover and respiration \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. In general, average assimilation rate per leaf area $\bar{p}$ increases with light level or canopy openness, $E$. The {\wplcp} can be estimated by solving for the the value of canopy openness $E=E^*$ giving $\frac{{\rm d}B} {{\rm d}t}(E)=0$ (Fig. \ref{fig:wplcp_idea}). From eqn \ref{eq:dbdt}, this occurs when
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum_{i={\rm l,b,s,r}} \, M_{\rm i} \left(\frac{k_{\rm i}}{y} + r_{\rm i}\right)}{A_{\rm l}}.
\end{equation}
The {\wplcp} occurs at the points where the photosynthetic production (per unit leaf area) line intersects with the sum of maintenance and respiration costs (per unit leaf area) for each tissue (Fig. \ref{fig:wplcp_idea}). Traits will influence the {\wplcp} if they effect either income or costs. Also, since the amount of stem support naturally increases with height, the {\wplcp} also naturally increases with height \citep{Givnish-1988} (Fig. \ref{fig:wplcp_idea}).

@@

It worth noting that because eqns \ref{eq:dMtdt}-\ref{eq:wplcp} are derived using standard rules of addition, multiplication and differentiation, and must therefore hold for any potential growth model where biomass allocation is important. To make explicit predictions via this framework, then requires an explicit model of plant construction and function; i.e. we must put equations behind all there terms in Fig. \ref{fig:conceptual}b.

The {\plant} adopts a model of plant construction and function that can be considered a first-order functional-balance or function-equilibrium model, similar to those implemented by \citet{Makela-1997} and \citet{Moorcroft-2001}. We could also call it ``isometric'', because the assumptions see area-based metrics scaling to the first-power of other area-based metrics, and to the square-power of length-based metrics \citep{Huxley-1932}. Table \ref{tab:allometry} provides key equations of the model (see \citealt{Falster-2016} for full derivation). In particular, we assume that as a plant grows:
\begin{description}
\item[Assumption 1:] Its height scales to the 0.5 power of its leaf area (eqn \ref{eq:ha}).
\item[Assumptions 2:] The cross-sectional area of sapwood in the stem is proportional to its leaf area (eqn \ref{eq:as}).
\item[Assumptions 3:] The cross-sectional area of bark and phloem in the stem  is proportional to its leaf area (eqn \ref{eq:ab}).
\item[Assumptions 4:] The cross-sectional area of root surface area and therefore mass is proportional to its leaf area (eqn \ref{eq:mr}).
\item[Assumption 5:] The vertical distribution of leaf within the plant's canopy, relative to the plant's height, remains constant.
\end{description}
Assumption 1 accounts for the architectural layout of the plant. Assumptions 2-4 are realisations of the pipe model \citep{Shinozaki-1964}, whereby the cross-sectional area of conducting tissues are proportional to leaf area. To describe the vertical distribution of leaf area within the canopy of an individual plant (assumption 5), we use the model of \citet{Yokozawa-1995}, which can account for a variety of canopy profiles through a single parameter $\eta_c$, varying from 0-1 (for details, see Supplementary Materials).

@@

@@

For the most part, parameters used in the current analysis were sourced from \citet{Falster-2016} (see Tables \ref{tab:params_core}--\ref{tab:params_hyper} for values). The only exceptions are: i) parameters effecting the relationships outlined in Table \ref{tab:allometry}a, which were fit using data described below, and ii) parameters describing the function for reproductive allocation. By default, we set $r_{\rm r1}=0.8$ and $r_{\rm r2}=10$, implying a relatively rapid transition to reproduction at $H=H_{{\rm mat}}$ (see panel for $\frac{{\rm d}M_{{\rm a}}}{{\rm d}B}$ in Fig. \ref{fig:conceptual}).

@@
